---
title: "Reproducible Research: Peer Assessment 1"
author: "Rich Robinson"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r init}
data_file <- "activity.csv"
zip_file <- "activity.zip"
zip_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

## Checking to see if required files are present
if (!file.exists(data_file)){
  
  if (!file.exists(zip_file)){
    download.file(zip_url, destfile = zip_file)
  }  
  unzip(zip_file)
}

act_data <- read.csv(data_file, header = TRUE, stringsAsFactors = FALSE)

```


## What is mean total number of steps taken per day?
```{r mean, message=FALSE, warning=FALSE}
library(dplyr)

res <- act_data %>% group_by(date) %>% summarise(total.steps=sum(steps))

avgs <- res %>% summarise(mean.steps=mean(total.steps, na.rm=TRUE), median.steps=median(total.steps, na.rm=TRUE))
```
```{r mean-plot, message=FALSE, warning=FALSE}
library(ggplot2)

base <- ggplot(res, aes(total.steps)) + geom_bar(fill="steelblue")
g <- base + labs(title = "Histogram of Total Steps per Day") + xlab("Number of Steps in a Day") + ylab("Frequency")
g <- g + theme(plot.title = element_text(face="bold"))

print(g)

```

The histogram above shows that the most common number of steps taken in an individual day is just above 10000.

The mean number of steps per day is `r avgs$mean.steps`, and the median number of steps per day is `r avgs$median.steps`.


## What is the average daily activity pattern?
```{r daily}
## Calculating the mean steps per intrval
ints <- act_data %>% group_by(interval) %>% summarise(mean.steps=mean(steps, na.rm=TRUE))

## Plotting
g2 <- ggplot(ints, aes(interval, mean.steps)) + geom_line(color="steelblue", size=1)
g2 <- g2 + labs(title="Average Number of Steps Across all Days per 5sec Interval") + theme(plot.title = element_text(face="bold"))
g2 <- g2 + xlab("Time in seconds") + ylab("Mean number of steps")

## Sorting the data in descending order by steps
ints2 <- ints %>% arrange(desc(mean.steps))

print(g2)

```

The grpah above shows how the average number of steps in a 5 second interval varies across a day. The values in each intervale were averaged across all days in the sample.

The interval with the highest average number of steps is intervale number `r ints2$interval[1]/5`, which is at `r ints2$interval[1]` seconds. The mean number of steps measured in this interval is `r ints2$mean.steps[1]`.

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
